State and Parameter Estimation for Spatio-Temporal Neural Fields

### 2010-05-08 ltx/Journal of Neuroengineering/StateAndParameterEstimationNeuralFields_JoNE.tex
periodic commit. making changes to discussion

The firing rate of the presynaptic neurons is related to the postsynaptic membrane potential by the sigmoidal activation function 
\begin{equation}
	\label{ActivationFunction} f\left( v\left( \mathbf{r}', t \right) \right) = \frac{f_{max}\left(\mathbf{r}\right)}{1 + \exp \left( \varsigma \left( v_0\left( \mathbf{r} \right) - v\left(\mathbf{r}',t\right) \right) \right)}. 
\end{equation}
The parameters $f_{max}$ and $v_0$ describe the maximum firing rate and firing threshold of the neural populations. The parameter $\varsigma$ governs the slope of the sigmoid. By substituting equation~\ref{RateBasedInteractions} into \ref{SpikesToPotential} we get the spatiotemporal model 
\begin{eqnarray}
	\label{FullDoubleIntModel} v\left(\mathbf{r},t\right) &=&  \int_{-\infty}^t h\left(t - t'\right) \\
	&\times&\int_\Omega w\left(\mathbf{r},\mathbf{r}'\right) f\left( v\left( \mathbf{r}',t \right)\right)d\mathbf{r}'dt'. \nonumber
\end{eqnarray}
To arrive at the final form of the model we shall state the synaptic response kernel as a Green's function 
\begin{equation}
	\label{GreensFuncDef} Dh\left( t \right) = \delta \left( t \right), 
\end{equation}
where $D=\frac{d}{dt} + \zeta$ is a temporal differential operator and $\delta(t)$ is the Dirac-delta function giving 
\begin{eqnarray}
	\label{FinalFormContinuous} \frac{dv\left( \mathbf{r},t \right)}{dt} &+& \zeta v\left( \mathbf{r},t \right) = \\
	&&\int_\Omega {w\left( \mathbf{r},\mathbf{r}' \right)f\left( {v\left( \mathbf{r}',t \right)} \right)dr'}. \nonumber
\end{eqnarray}
To arrive at the integro-difference equation (IDE) form of the model we discretise time using a first-order Euler method (see Appendix~\ref{Time Discretization}) giving 
\begin{eqnarray}
	\label{DiscreteTimeModel} v_{t+1}\left(\mathbf{r}\right) &=& \xi v_t\left(\mathbf{r}\right) + T_s \int_\Omega { w\left(\mathbf{r}-\mathbf{r}'\right) f\left(v_t\left(\mathbf{r}'\right)\right) d\mathbf{r}'} \nonumber\\ 
	&+& e_t\left(\mathbf{r}\right), 
\end{eqnarray}
where $T_s$ is the time step, $\xi = 1-\zeta T_s$ and $e_t(\mathbf{r})$ is i.i.d. such that $e_t\sim\mathcal{GP}(\mathbf 0,\gamma(\mathbf{s}-\mathbf{r}))$. This term was added to account for uncertainty in the model. To generate iEEG/LFP data we used the output function 
\begin{equation}
	\mathbf{y}_t = \int_{\Omega}{m\left(\mathbf{r}_n-\mathbf{r}'\right)v_t\left(\mathbf{r}'\right)d\mathbf{r}'} + \boldsymbol{\varepsilon}_t, 
\end{equation}
where $\mathbf{r}_n$ defines the location of the sensors in the field where $n=1,...,N$ indexes the sensors and $\boldsymbol{\varepsilon}_t \sim \mathcal{N}\left(0,\Sigma_{\varepsilon}\right)$. The output kernel $m(\mathbf{r}-\mathbf{r}')$ governs the geometry of pick-up range of sensors where 
\begin{equation}
	m\left(\mathbf{r}-\mathbf{r}'\right) = \exp{\left(-\frac{(\mathbf{r}-\mathbf{r}')^\top(\mathbf{r}-\mathbf{r}')}{\sigma_m^2}\right)}. 
\end{equation}

@@

\begin{figure*}[!th] 
\subfigure[][]{\includegraphics[width=2in]{./Graph/TrueField90.pdf}}
\subfigure[][]{\includegraphics[width=2in]{./Graph/EstimatedField90.pdf}}
\subfigure[][]{\includegraphics[width=2in]{./Graph/EstimationError90.pdf}}\\
\caption{One snapshot of the spatial field; a: True field; b: Estimated field; c: The absolute error.}
\label{fig:FieldEstimate}
\end{figure*}
\section{Discussion}\label{DiscussionSection}

@@

\begin{itemize}
	\item if we don't know the frequency analysis or the system is marginally sampled or under-sampled, the best we can do is have the basis functions the same as the sensor kernel. \wtf{why wouldn't it be better to have, say, twice as many basis functions as we have sensors, at half the width?} 
	\item alternatives to frequency analysis are model selection... \wtf{This frequency analysis \emph{is} model selection. Maybe what's meant here is that other model selection approaches could be applied? Some sort of IC optimisation?} 
	\item the width of the connectivity basis functions \wtf{do you mean width of the kernel? I'm not sure about this - as the parameters increase the kernel width increases right? So maybe you mean the basis functions, but don't we estimate these widths from the frequency analysis?} here is known exactly. In practice we don't 
	\item the number of basis functions could be set higher to get a better representation of the field, however this is limited by computation complexity. \wtf{Unless you are going to specify the computational complexity, in terms of the number of basis functions, it might be better to say something like ``there is a relationship between the number of basis functions that represent the field and the computational complexity of the estimation algorithm - the more basis functions the longer it takes to estimate''} 
	\item Discuss the novelty of using the reduced model (BrainIDE) for creating patient specific models. \omg{hell yes. This should be front and centre!} 
	\item The size of the dentritic arbour is consistent in the cortex. \omg{?} 
	\item Spatial frequency of field and the relationship to the required density of electrodes/sensors required for estimation or to capture the required information without aliasing. 
	\item The flexibility of the approach 
	\item Combining this method with other to incorporate longer range hetrogeneous connectivity. \omg{awesome} 
	\item The need for activity to perform estimation. \omg{To Learn You Must Perturb!} 
	\item Recent studies using neural field models have demonstrated the theoretical implications of specific connectivity structure of the cortical columns in neural field models, where the connectivity kernel governs the bifurcation points of the system \cite{Hutt2005} and types of oscillations that can be generated \cite{Schmidt2009}. This implies that if we could estimate the connectivity structure for an individual, then we could capture patient specific neurodynamics that lead to various oscillatory states.
	\item Estimating functional cortical connectivity via EEG promises to shed light on the hierarchy of brain regions that are involved in specific tasks, motivating the use of techniques like DCM. Other comparable methods are based on auto-regressive (AR) modelling of the EEG, MEG, or fMRI times series \todo{[refs!]}. Studies have concentrated on finding functional connectivity patterns using information contained in the AR coefficients using techniques like Granger causality \cite{Hesse2003}, the direct transfer function \cite{Kaminski1991}, and partial directed coherence \cite{Sameshima1999}. Again, like the DCM approach, these methods estimate long range connectivity patterns that does not provide a clear relationship between the continuum field models and data.
	\item Long range connectivity can be estimated using DTI \cite{Knock2009}, however this short range connectivity can not estimated using DTI
\end{itemize}

@@

& -&\zeta v\left( \mathbf{r},t \right) + \int_\Omega {w\left( \mathbf{r}-\mathbf{r}' \right)f\left( {v\left( \mathbf{r}',t \right)} \right)d\mathbf{r}'} + u\left(\mathbf{r},t\right).\nonumber \\ 

@@

&+& T_s \int_\Omega {w\left( \mathbf{r}-\mathbf{r}' \right)f\left( {v_t\left( \mathbf{r}'\right)} \right)d\mathbf{r}'} + T_s u_t\left(\mathbf{r}\right).\nonumber \\ 

@@

	\label{Discrete Time Model1} v_{t+1}\left(\mathbf{r}\right) = \xi v_t\left(\mathbf{r}\right) + T_s \int_\Omega { w\left(\mathbf{r}-\mathbf{r}'\right) f\left(v_t\left(\mathbf{r}'\right)\right) d\mathbf{r}'} + T_s u_t\left(\mathbf{r}\right), 

@@

&+& T_s u_t\left(\mathbf{r}\right) + e_t(\mathbf{r}_{i,j}), 

@@

&-& \xi\int_\Omega {\boldsymbol{\phi}(\mathbf{r})\boldsymbol{\phi}^{\top}(\mathbf{r})d\mathbf{r}} \mathbf{x}_t+T_s \int_\Omega{\boldsymbol{\phi} \left(\mathbf{r}\right) u_t\left(\mathbf{r}\right)d\mathbf{r}}\nonumber\\
&+&\int_\Omega{\boldsymbol{\phi} \left(\mathbf{r}\right) e_t\left(\mathbf{r}\right)d\mathbf{r}}. 

@@

Substituting Eq.~\ref{DefGamma} into Eq.~\ref{StartofReduction} and cross-multiplying by $\boldsymbol{\Gamma}^{-1}$ gives 

@@

	 \lefteqn{\mathbf{x}_{t+1} \nonumber= T_s\boldsymbol{\Gamma}^{ - 1}\int_\Omega {\boldsymbol{\phi}(\mathbf{r}) \int_\Omega {f(\boldsymbol{\phi}^{\top}(\mathbf{r}')\mathbf{x}_t) \boldsymbol{\psi}^{\top} (\mathbf{r}-\mathbf{r}')d\mathbf{r}'} d\mathbf{r}} \boldsymbol{\theta}} \nonumber\\
&-&\xi\mathbf{x}_t+\boldsymbol{\Gamma}^{-1}T_s\int_\Omega{\boldsymbol{\phi}(\mathbf{r}) u_t(\mathbf{r})d\mathbf{r}}\nonumber+\boldsymbol{\Gamma}^{-1} \int_\Omega{\boldsymbol{\phi}(\mathbf{r})e_t(\mathbf{r})d\mathbf{r}}.\\ 

>>>>

The firing rate of the presynaptic neurons is related to the postsynaptic membrane potential by the sigmoidal activation function 
\begin{equation}
	\label{ActivationFunction} f\left( v\left( \mathbf{r}', t \right) \right) = \frac{f_{max}\left(\mathbf{r}\right)}{1 + \exp \left( \varsigma \left( v_0\left( \mathbf{r} \right) - v\left(\mathbf{r}',t\right) \right) \right)}. 
\end{equation}
The parameters $f_{max}$ and $v_0$ describe the maximum firing rate and firing threshold of the neural populations. The parameter $\varsigma$ governs the slope of the sigmoid. By substituting equation~\ref{RateBasedInteractions} into \ref{SpikesToPotential} we get the spatiotemporal model 
\begin{eqnarray}
	\label{FullDoubleIntModel} v\left(\mathbf{r},t\right) &=&  \int_{-\infty}^t h\left(t - t'\right) \\
	&\times&\int_\Omega w\left(\mathbf{r},\mathbf{r}'\right) f\left( v\left( \mathbf{r}',t \right)\right)d\mathbf{r}'dt'. \nonumber
\end{eqnarray}
To arrive at the final form of the model we shall state the synaptic response kernel as a Green's function 
\begin{equation}
	\label{GreensFuncDef} Dh\left( t \right) = \delta \left( t \right), 
\end{equation}
where $D=\frac{d}{dt} + \zeta$ is a temporal differential operator and $\delta(t)$ is the Dirac-delta function giving 
\begin{eqnarray}
	\label{FinalFormContinuous} \frac{dv\left( \mathbf{r},t \right)}{dt} &+& \zeta v\left( \mathbf{r},t \right) = \\
	&&\int_\Omega {w\left( \mathbf{r},\mathbf{r}' \right)f\left( {v\left( \mathbf{r}',t \right)} \right)d\mathbf{r}'}. \nonumber
\end{eqnarray}
To arrive at the integro-difference equation (IDE) form of the model we discretise time using a first-order Euler method (see Appendix~\ref{Time Discretization}) giving 
\begin{eqnarray}
	\label{DiscreteTimeModel} v_{t+1}\left(\mathbf{r}\right) &=& \xi v_t\left(\mathbf{r}\right) + T_s \int_\Omega { w\left(\mathbf{r}-\mathbf{r}'\right) f\left(v_t\left(\mathbf{r}'\right)\right) d\mathbf{r}'} \nonumber\\ 
	&+& e_t\left(\mathbf{r}\right), 
\end{eqnarray}
where $T_s$ is the time step, $\xi = 1-\zeta T_s$ and $e_t(\mathbf{r})$ is i.i.d. such that $e_t\sim\mathcal{GP}(\mathbf 0,\gamma(\mathbf{s}-\mathbf{r}))$. This term was added to account for uncertainty in the model. To generate iEEG/LFP data we used the output function 
\begin{equation}
	\mathbf{y}_t = \int_{\Omega}{m\left(\mathbf{r}_n-\mathbf{r}'\right)v_t\left(\mathbf{r}'\right)d\mathbf{r}'} + \boldsymbol{\varepsilon}_t, 
\end{equation}
where $\mathbf{r}_n$ defines the location of the sensors in the field where $n=1,...,N$ indexes the sensors and $\boldsymbol{\varepsilon}_t \sim \mathcal{N}\left(0,\Sigma_{\varepsilon}\right)$. The output kernel $m(\mathbf{r}-\mathbf{r}')$ governs the geometry of pick-up range of sensors where 
\begin{equation}
	m\left(\mathbf{r}-\mathbf{r}'\right) = \exp{\left(-\frac{(\mathbf{r}-\mathbf{r}')^\top(\mathbf{r}-\mathbf{r}')}{\sigma_m^2}\right)}. 
\end{equation}

@@

\begin{figure*}[!th] 
\subfigure[][]{\includegraphics[width=2in]{./Graph/TrueField90.pdf}}
\subfigure[][]{\includegraphics[width=2in]{./Graph/EstimatedField90.pdf}}
\subfigure[][]{\includegraphics[width=2in]{./Graph/EstimationError90.pdf}}\\
\caption{One snapshot of the spatial field; a: True field; b: Estimated field; c: The absolute error.}
\label{fig:FieldEstimate}
\end{figure*}

\section{Discussion}\label{DiscussionSection}
This paper presents a novel framework for model reduction to make estimation of neural field practical. This systematic model reduction allows for the choice a complexity. The level of detail of the field is related to the number and width of the field basis functions, thereby providing a flexible approach where detail and computational complexity can be tradeoff. For certain application it might be sufficient to capture only low (spatial) frequency dynamics.

The frequency analysis and model selection tools presented in this paper specifies the sensor and field basis function arrangement required to form a state-space model to capture and model dominant cortical dynamics. These tools can be used to better design experiments and recording techniques to advance model based neural parameter estimation and avoid spatial aliasing. In addition, the assumption homogeneity can be relaxed where heterogeneity can be model by careful placement of field basis functions. For example, if cortical regions have a greater density of neural populations, then more basis functions can be used in these areas. In this scenario, model selection and basis function positioning can be achieved using instantaneous frequency analysis of the the neural field or alternative model selection tools based on information criteria.

To the authors best knowledge, this is the first paper to propose an estimation framework for intracortical connectivity from electrophysiological data. This is important as the connectivity structure at this scale can not be estimated using diffusion tensor imaging due to the fine spatial scale and the high degree of crossing of fibres\omg{ref}. In addition, other non-model based techniques for estimating functional connectivity such as Granger causality \cite{Hesse2003}, the direct transfer function \cite{Kaminski1991}, and partial directed coherence \cite{Sameshima1999} are restricted to the spatial scales of the recording electrodes. The model based technique proposed in this paper estimates the connectivity structure independently of the recording electrode geometry. An advantage of these auto-regressive (AR) based methods is that they allow for anisotropic connectivity, which is present in cortico-cortical fibres. The AR methods are complementary to the proposed model based framework and by combining these methods a more realistic model can be estimated. A recent theoretical study~\cite{Jirsa2009} has demonstrated the importance of this multi-scale structure in generating the characteristic rhythms of the brain.

Long range connectivity can be estimated using DTI \cite{Knock2009}, however this short range connectivity can not estimated using DTI

@@

This estimation framework also has theoretical implications. This estimation framework has the potential to allow testing hypotheses that have been generated in theoretical studies and to validate neural field models. For example, it has been hypothesised that so called bump solutions are a possible mechanism for short term memory formation~\cite{Coombes2005}. Using this framework, we can reconstruct a neural field from data, and estimate parameters and check if they correspond the the theoretically derived parameters where these bump solutions can exist. Therefore, this framework has the potential to validate models and test theoretically derived hypotheses. Associating regions of state-space and parameter sets to cognitive can potentially validate theoretical findings. For example, sleep, wake, memory,

Here we assumed knowledge of the width of the kernel basis functions. If a future work we plan to extend this work to unknown widths.

In future work, we aim to show how heterogeneous can be modelled within this formulation. We will still assume a isotropic connectivity kernel however heterogeneities can be modelled by placing basis function c

\begin{itemize}
	\item if we don't know the frequency analysis or the system is marginally sampled or under-sampled, the best we can do is have the basis functions the same as the sensor kernel. \wtf{why wouldn't it be better to have, say, twice as many basis functions as we have sensors, at half the width?} 
	\item alternatives to frequency analysis are model selection... \wtf{This frequency analysis \emph{is} model selection. Maybe what's meant here is that other model selection approaches could be applied? Some sort of IC optimisation?}

	\item The size of the dentritic arbour is consistent in the cortex. \omg{?}

	\item The need for activity to perform estimation. \omg{To Learn You Must Perturb!} 
	\item Recent studies using neural field models have demonstrated the theoretical implications of specific connectivity structure of the cortical columns in neural field models, where the connectivity kernel governs the bifurcation points of the system \cite{Hutt2005} and types of oscillations that can be generated \cite{Schmidt2009}. This implies that if we could estimate the connectivity structure for an individual, then we could capture patient specific neurodynamics that lead to various oscillatory states.
\end{itemize}

@@

& -&\zeta v\left( \mathbf{r},t \right) + \int_\Omega {w\left( \mathbf{r}-\mathbf{r}' \right)f\left( {v\left( \mathbf{r}',t \right)} \right)d\mathbf{r}'}.\nonumber \\ 

@@

&+& T_s \int_\Omega {w\left( \mathbf{r}-\mathbf{r}' \right)f\left( {v_t\left( \mathbf{r}'\right)} \right)d\mathbf{r}'}.\nonumber \\ 

@@

	\label{Discrete Time Model1} v_{t+1}\left(\mathbf{r}\right) = \xi v_t\left(\mathbf{r}\right) + T_s \int_\Omega { w\left(\mathbf{r}-\mathbf{r}'\right) f\left(v_t\left(\mathbf{r}'\right)\right) d\mathbf{r}'}, 

@@

&+& e_t(\mathbf{r}_{i,j}), 

@@

&-& \xi\int_\Omega {\boldsymbol{\phi}(\mathbf{r})\boldsymbol{\phi}^{\top}(\mathbf{r})d\mathbf{r}} \mathbf{x}_t+
\int_\Omega{\boldsymbol{\phi} \left(\mathbf{r}\right) e_t\left(\mathbf{r}\right)d\mathbf{r}}. 

@@

Substituting equation~\ref{DefGamma} into equation~\ref{StartofReduction} and cross-multiplying by $\boldsymbol{\Gamma}^{-1}$ gives 

@@

	 \lefteqn{\mathbf{x}_{t+1} \nonumber = T_s\boldsymbol{\Gamma}^{ - 1}\int_\Omega {\boldsymbol{\phi}(\mathbf{r}) \int_\Omega {f(\boldsymbol{\phi}^{\top}(\mathbf{r}')\mathbf{x}_t) \boldsymbol{\psi}^{\top} (\mathbf{r}-\mathbf{r}')d\mathbf{r}'} d\mathbf{r}} \boldsymbol{\theta}} \nonumber\\
&-&\xi\mathbf{x}_t + \boldsymbol{\Gamma}^{-1} \int_\Omega{\boldsymbol{\phi}(\mathbf{r})e_t(\mathbf{r})d\mathbf{r}}.\\ 
